---
title: "Lending club"
author: "Samir Gadkari"
date: "3/24/2021"
output: html_document
---

```{r include=FALSE}
library(tidyverse)
library(readr)
library(lubridate)
```

## 1.1 Read in data
Since the accepted/rejected files are large, grep it and save all "credit_card" rejections to another file. This will make it faster (and possible) to load it in memory during the read_csv call.

```{bash}
FILE=$DATASETS/financial_datasets/all_lending_club_loan_data/accepted_credit_card_2007_to_2018Q4_2.csv

if [ ! -f "$FILE" ]; then
  grep "credit_card" $DATASETS/financial_datasets/all_lending_club_loan_data/accepted_2007_to_2018Q4_2.csv >> $DATASETS/financial_datasets/all_lending_club_loan_data/accepted_credit_card_2007_to_2018Q4_2.csv
fi

FILE=$DATASETS/financial_datasets/all_lending_club_loan_data/rejected_credit_card_2007_to_2018Q4_2.csv

if [ ! -f "$FILE" ]; then
  grep "credit_card" $DATASETS/financial_datasets/all_lending_club_loan_data/rejected_2007_to_2018Q4_2.csv >> $DATASETS/financial_datasets/all_lending_club_loan_data/rejected_credit_card_2007_to_2018Q4_2.csv
fi
```

```{r}
data_path <- paste0(Sys.getenv("DATASETS"), "/financial_datasets",
                   "/all_lending_club_loan_data")

accepted_cc <- read_csv(
  paste0(data_path, "/accepted_credit_card_2007_to_2018Q4_2.csv"),
  col_types = cols(desc = col_character(),
                   revol_bal_joint = col_double(),
                   sec_app_fico_range_low = col_double(), 
                   sec_app_fico_range_high = col_double(),
                   sec_app_earliest_cr_line =
                     col_double(),    
                   sec_app_inq_last_6mths = col_double(),
                   sec_app_mort_acc = col_double(),
                   sec_app_open_acc = col_double(),
                   sec_app_revol_util = col_double(),
                   sec_app_open_act_il = col_double(),    
                   sec_app_num_rev_accts = col_double(),
                   sec_app_earliest_cr_line = 
                     col_skip(),
                   sec_app_chargeoff_within_12_mths =
                     col_double(),
                   sec_app_collections_12_mths_ex_med =
                     col_double(),
                   sec_app_mths_since_last_major_derog =
                     col_double()
                  ),
  progress = show_progress())

head(accepted_cc)
nrow(accepted_cc)
```
```{r}
nrow(accepted_cc[accepted_cc$purpose == "credit_card", ])
```

Great !! Now we have half a million row values to work with (and that's just for the accepted credit applications).

## 1.2 Select columns and rows that are not full of NAs

Let's see which columns are not full of NAs.
```{r}
(selected_cols <- accepted_cc %>%
  select(which(colMeans(is.na(.)) < 0.02)) %>%
  names())
```
Now we have the columns where the number of NAs are less than 2%. We should do this, since this is lending club data, and columns which are expected to be useful will not have significant NAs.

Let's select these columns, and remove any rows that have NA values in it. This way, we won't have to deal with any missing values. If our model is bad, we can always interpolate the missing values.

```{r}
accepted_cc <- na.omit(accepted_cc[ , selected_cols])
head(accepted_cc)
```

Now we have a large number of rows, enough to work with.

```{r}
print(paste(names(accepted_cc), collapse = ",  "))
```
## 1.3 List unique values in character columns

```{r}
unique_values <- function(ref) {
  paste(unique(ref), collapse = ",   ")
}

unique_char_col_vals <- function(df_name, df) {
  writeLines(
    stringr::str_glue("{df_name}: ------------------------------->>>>>"))
  
  result <- as.list(df %>%
        select(df %>%
                 select_if(is.character) %>%
                 select(!contains("date", ignore.case = TRUE)) %>%
                 summarise(across(everything(), n_distinct)) %>%
                 select(where(~.x[[1]] < 40)) %>%
                 names()) %>%
        summarise(across(everything(), unique_values)))

  print(result)
  
  writeLines(
    stringr::str_glue("<<<<<--------------------------- :{df_name}"))
}

unique_char_col_vals("accepted_cc", accepted_cc)
```
## 1.4 Understanding the id column

```{r}
n_occur <- data.frame(table(as.integer(accepted_cc$id)))
n_occur[n_occur$Freq > 1, ]
remove(n_occur)
```


Each id value is unique, and none of them are member IDs. This means each line is for a separate loan.

```{r}
print(paste(names(accepted_cc), collapse = ",  "))
```

## 1.5 Date columns

### 1.5.1 Convert date character columns to POSIXct format

Convert all date columns to POSIXct for ease of use later. There is no day specified, and POSIXct requires it. We will add the first day of the month to each date.

```{r}
to_day_month_year <- function(month_year) {
  as.POSIXct(strptime(paste0("01-", month_year), 
                      "%d-%b-%Y"))
}

accepted_cc <- accepted_cc %>%
  mutate(
    issue_date = to_day_month_year(issue_d), .after = issue_d
  ) %>%
  mutate(
    earliest_cr_line = to_day_month_year(earliest_cr_line),
    last_pymnt_d = to_day_month_year(last_pymnt_d),
    last_credit_pull_d = 
      to_day_month_year(last_credit_pull_d)
  ) %>%
  select(-issue_d)

min(accepted_cc[["issue_date"]])
max(accepted_cc[["issue_date"]])
```

There are accounts from `min(accepted_cc[["issue_date"]]` through `min(accepted_cc[["issue_date"]]`.

### 1.5.2 Distribution of date columns

Let's look at the distribution of the dates:

```{r}
accepted_cc %>%
  ggplot(aes(x = issue_date)) +
  geom_histogram(bins = 30)

accepted_cc %>%
  ggplot(aes(x = earliest_cr_line)) +
  geom_histogram(bins = 30)
```

## 1.6 Numerical columns

```{r}
table(accepted_cc$pub_rec)              # public records count?
table(accepted_cc$pub_rec_bankruptcies) # public record bankruptcies count
```

> We may need these columns. There are many bankruptcies which we can use
> to filter out bad applicants - especially when there are more than one.

```{r}
sum(!near(accepted_cc$funded_amnt,
          accepted_cc$funded_amnt_inv, tol = 700.0))
head(accepted_cc[!near(accepted_cc$funded_amnt,
                       accepted_cc$funded_amnt_inv, tol = 700.0), 
                 c("funded_amnt", "funded_amnt_inv")])
sum(!near(accepted_cc$loan_amnt,
          accepted_cc$funded_amnt, tol = 700.0))
head(accepted_cc[!near(accepted_cc$loan_amnt,
                       accepted_cc$funded_amnt, tol = 700.0),
                 c("loan_amnt", "funded_amnt")])
```

> The funded amounts, and the loan amounts are close to each other for almost 
> all of the data.
> Since there is very little information in these three columns together,
> we can remove two of them. Let's keep the funded amount column.

```{r}
accepted_cc <- accepted_cc %>% 
  select(-c(loan_amnt, funded_amnt_inv))
```


```{r}
sum(!near(accepted_cc$total_pymnt, 
          accepted_cc$total_pymnt_inv, tol = 700.0))
head(accepted_cc[!near(accepted_cc$total_pymnt, 
          accepted_cc$total_pymnt_inv, tol = 700.0), 
          c("total_pymnt", "total_pymnt_inv")])
```

> The payment amounts are close to each other for almost all of the data.
> Since there is very little information in these two columns together,
> we can remove one of them.

```{r}
accepted_cc <- accepted_cc %>%
  select(-total_pymnt_inv)
```

```{r}
sum(!near(accepted_cc$out_prncp,
          accepted_cc$out_prncp_inv, tol = 700.0))
```

> The two columns (out_prncp, out_prncp_inv) are almost the same.
> We will remove one of them

```{r}
accepted_cc <- accepted_cc %>%
  select(-out_prncp_inv)

accepted_cc %>%
  select(funded_amnt, term, int_rate, installment,
         revol_bal, revol_util, total_pymnt, total_rec_prncp, out_prncp, total_rec_int,
         last_pymnt_amnt, last_pymnt_d, bc_open_to_buy, bc_util, mths_since_recent_bc,
         num_bc_sats, num_sats, percent_bc_gt_75, total_bal_ex_mort,
         total_bc_limit)
```


```{r}
unique(accepted_cc$policy_code)
```

> Only one policy code - so remove the column.
> 
> The url column is not useful - it's been a while, and none of the urls work.
> Remove the column.
> The purpose and title columns are also not useful. The stated purpose is
> for credit card refinancing/payments, but there is no guarantee that is what
> the loan is going to be used for.

```{r}
accepted_cc <- accepted_cc %>%
  select(-c(policy_code, url, purpose, title))
```

```{r}
head(accepted_cc[accepted_cc$total_bc_limit, ])
```

```{r}
table(accepted_cc$tax_liens)
table(accepted_cc$debt_settlement_flag)
table(accepted_cc$hardship_flag)
```

> There are very few hardship cases - remove them.

```{r}
table(accepted_cc$hardship_flag)
accepted_cc <- accepted_cc %>%
  filter(hardship_flag == "N")

table(accepted_cc$hardship_flag)
accepted_cc <- accepted_cc %>%
  select(-c(hardship_flag))
```

```{r}
table(accepted_cc$inq_last_6mths)
```

```{r}
table(accepted_cc$delinq_2yrs)
```
```{r}
table(accepted_cc$pymnt_plan)
```

> There are only 96 payment plans set up. We can ignore this column,
> or better yet, remove these rows, as they will get averaged out.

```{r}
if ("y" %in% accepted_cc$pymnt_plan) {
  accepted_cc <- accepted_cc %>%
    filter(pymnt_plan == "y")
}

table(accepted_cc$pymnt_plan)

accepted_cc <- accepted_cc %>%
  select(-c(pymnt_plan))
```


```{r}
table(accepted_cc$verification_status)
```

```{r}
table(accepted_cc$home_ownership)
```

> We can remove the ANY, NONE, OTHER rows.

```{r}
accepted_cc <- accepted_cc %>%
  filter(home_ownership %in% c("MORTGAGE", "OWN", "RENT"))
table(accepted_cc$home_ownership)
```

```{r}
table(accepted_cc$term)
```
```{r}
table(accepted_cc$mort_acc)
```

```{r}
table(accepted_cc$chargeoff_within_12_mths)
```

```{r}
head(accepted_cc$installment)
```

```{r}
table(accepted_cc$acc_now_delinq)
```

```{r}
table(accepted_cc$acc_open_past_24mths)
```

```{r}
head(accepted_cc)
```

## 1.6.1 Remove numerical columns that are not useful

```{r}
accepted_cc <- accepted_cc %>%
  select(-c("mths_since_recent_bc", "num_bc_sats", "num_sats", 
            "percent_bc_gt_75", "tax_liens", "total_bal_ex_mort", 
            "total_bc_limit", "disbursement_method"))
```

```{r}
table(accepted_cc$loan_status)
```

```{r}
print(paste0(names(accepted_cc), collapse = ", "))
accepted_cc
```

## 1.7 Split the data into training, validation, test sets

There is not enough data to split by term of the loan (36-month, 60-month), and apply the processing to each term type. When this is done, there are some cohorts with NAs in some of the loan_status. This is why we will go with the full dataset.

Keep 60% in the training dataset, 20% in the validation dataset, and 20% in the testing dataset. For inference, the training dataset is used. For prediction, the training and validation datasets are used. Once satisfied, the testing dataset is used only once to get the final prediction results.

```{r}
accepted_cc <- accepted_cc %>%
  mutate(n = row_number()) %>% # create row number column
  select(n, everything())

train <- accepted_cc %>%
  group_by(loan_status) %>%
  sample_frac(0.6)

validate_test <- accepted_cc %>% anti_join(train)

validate <- validate_test %>%
  group_by(loan_status) %>%
  sample_frac(0.5)

test <- validate_test %>% anti_join(validate)
remove(validate_test)
```

```{r}
nrow(accepted_cc)
nrow(train)
nrow(validate)
nrow(test)
```

```{r}
table(train$loan_status)
table(validate$loan_status)
table(test$loan_status)
```

It's good - each dataset has at least 1 value of each attribute of loan_status.

## 1.8 Get cohorts using the training set

We're analyzing cohorts, so we will split the data into multiple overlapping cohorts. Aggregating those cohorts will allow us to make predictions over business cycles.

```{r}
range(train$issue_date)
```

The range is 6 years 6 months. To split it into 3, we will give each cohort 2 years 2 months.

```{r}
min_cohort_date <- min(train$issue_date)
max_cohort_date <- max(train$issue_date)

cohorts <- list(c(min_cohort_date, 
                  add_with_rollback(
                    min_cohort_date, period(years = 2, months = 2))
                  ),
                c(add_with_rollback(
                    min_cohort_date, period(years = 2, months = 2)),
                  add_with_rollback(
                    min_cohort_date, period(years = 4, months = 4))
                  ),
                c(add_with_rollback(
                    min_cohort_date, period(years = 4, months = 4)),
                  add_with_rollback(
                    min_cohort_date, period(years = 6, months = 6)))
                )

cohorts

# We assert that the end dates match
testthat::expect_equal(cohorts[[1]][[1]], min_cohort_date)
testthat::expect_equal(cohorts[[3]][[2]], max_cohort_date)
```

```{r}
train <- train %>%
  mutate(
    cohort = case_when(
      issue_date %within% interval(cohorts[[1]][[1]], 
                                   cohorts[[1]][[2]]) ~ 1,
      issue_date %within% interval(cohorts[[2]][[1]], 
                                   cohorts[[2]][[2]]) ~ 2,
      issue_date %within% interval(cohorts[[3]][[1]], 
                                   cohorts[[3]][[2]]) ~ 3
      )
  ) 

train %>%
  group_by(cohort) %>%
  summarise(
    n = n()
  )

train %>%
  ggplot(aes(issue_date)) +
    geom_histogram(bins = 30) +
    facet_wrap(~cohort)
```

```{r}
cohort_groupings <- train %>%
  group_by(cohort) %>%
  summarise(
    names = names(table(loan_status)),
    num_rows = table(loan_status))

cohort_groupings <- 
  pivot_wider(cohort_groupings, names_from = names,
              values_from = num_rows)
cohort_groupings
```

This is what the [Lending Club site](https://help.lendingclub.com/hc/en-us/articles/215488038-What-do-the-different-Note-statuses-mean-) says about the meaning of each loan status:

  * Current: Loan is up to date on all outstanding payments. 
  * In Grace Period: Loan is past due but within the 15-day grace period.
  * Late (16-30): Loan has not been current for 16 to 30 days. Learn more about the
    tools LendingClub has to deal with delinquent borrowers.
  * Late (31-120): Loan has not been current for 31 to 120 days. Learn more about the
    tools LendingClub has to deal with delinquent borrowers.
  * Fully paid: Loan has been fully repaid, either at the expiration of the 3- or
    5-year year term or as a result of a prepayment.
  * Default: Loan has not been current for an extended period of time. Learn more
    about the difference between “default” and “charge off”. 
  * Charged Off: Loan for which there is no longer a reasonable expectation of
    further payments. Charge Off typically occurs when a loan is 120 days or more
    past due. 

So the loan status follows this path:
Current -> In Grace Period -> Late (16-30 days) -> Late (31-120 days) -> Charged Off
                                                        Default
Default occurs before Charged Off, but is not specified precisely where.
Fully Paid can occur before the last payment (if pre-payments are made), or at the time of the last payment, or some time later when the borrower has enough funds.

We should convert the loan_status column into a DPD (days past due) column. We can use this mapping:

  * fully paid: (-Inf, -1) dpd. Numbers are given in a (min, max) range.
  * current: (0, 0) dpd
  * in grace period: (1, 15) dpd
  * late (16-30 days): (16, 30) dpd
  * late (31-120 days): (31, 120) dpd
  * default: (45, 60) dpd
  * charged off: (121, Inf) dpd

```{r}
train <- train %>%
  mutate(
    dpd_min = case_when(
      loan_status == "Fully Paid" ~ -Inf,
      loan_status == "Current" ~ 0,
      loan_status == "In Grace Period" ~ 1,
      loan_status == "Late (16-30 days)" ~ 16,
      loan_status == "Late (31-120 days" ~ 31,
      loan_status == "Default" ~ 45,
      loan_status == "Charged Off" ~ 121
    ),
    dpd_max = case_when(
      loan_status == "Fully Paid" ~ -1,
      loan_status == "Current" ~ 0,
      loan_status == "In Grace Period" ~ 15,
      loan_status == "Late (16-30 days)" ~ 30,
      loan_status == "Late (31-120 days" ~ 120,
      loan_status == "Default" ~ 60,
      loan_status == "Charged Off" ~ Inf
    ),
    .after = loan_status
  )

head(train)
```

