---
title: "Lending club"
author: "Samir Gadkari"
date: "3/24/2021"
output: html_document
---

```{r include=FALSE}
library(tidyverse)
library(readr)
```

## 1.1 Read in data
Since the accepted/rejected files are large, grep it and save all
"credit_card" rejections to another file. This will make it faster
(and possible) to load it in memory during the read_csv call.

```{bash}
grep "id,member_id," $DATASETS/financial_datasets/all_lending_club_loan_data/accepted_2007_to_2018Q4_2.csv > $DATASETS/financial_datasets/all_lending_club_loan_data/accepted_credit_card_2007_to_2018Q4_2.csv

grep "credit_card" $DATASETS/financial_datasets/all_lending_club_loan_data/accepted_2007_to_2018Q4_2.csv >> $DATASETS/financial_datasets/all_lending_club_loan_data/accepted_credit_card_2007_to_2018Q4_2.csv

grep "Amount Requested,Application Date" $DATASETS/financial_datasets/all_lending_club_loan_data/rejected_2007_to_2018Q4_2.csv > $DATASETS/financial_datasets/all_lending_club_loan_data/rejected_credit_card_2007_to_2018Q4_2.csv

grep "credit_card" $DATASETS/financial_datasets/all_lending_club_loan_data/rejected_2007_to_2018Q4_2.csv >> $DATASETS/financial_datasets/all_lending_club_loan_data/rejected_credit_card_2007_to_2018Q4_2.csv
```

```{r}
data_path <- paste0(Sys.getenv("DATASETS"), "/financial_datasets",
                   "/all_lending_club_loan_data")

accepted_cc <- read_csv(
  paste0(data_path, "/accepted_credit_card_2007_to_2018Q4_2.csv"),
  col_types = cols(desc = col_character(),
                   revol_bal_joint = col_double(),
                   sec_app_fico_range_low = col_double(), 
                   sec_app_fico_range_high = col_double(),
                   sec_app_earliest_cr_line =
                     col_double(),    
                   sec_app_inq_last_6mths = col_double(),
                   sec_app_mort_acc = col_double(),
                   sec_app_open_acc = col_double(),
                   sec_app_revol_util = col_double(),
                   sec_app_open_act_il = col_double(),    
                   sec_app_num_rev_accts = col_double(),
                   sec_app_earliest_cr_line = 
                     col_skip(),
                   sec_app_chargeoff_within_12_mths =
                     col_double(),
                   sec_app_collections_12_mths_ex_med =
                     col_double(),
                   sec_app_mths_since_last_major_derog =
                     col_double()
                  ),
  progress = show_progress())

head(accepted_cc)
nrow(accepted_cc)
```
```{r}
nrow(accepted_cc[accepted_cc$purpose == "credit_card", ])
```

Maybe we can use the loan_status to show good vs bad loans.

Great !! Now we have half a million row values to work with (and that's
just for the accepted credit applications).
```{r}
print(paste(names(accepted_cc), collapse = ",  "))
```
```{r}
unique_values <- function(ref) {
  paste(unique(ref), collapse = ",   ")
}

unique_char_col_vals <- function(df_name, df) {
  writeLines(
    stringr::str_glue("{df_name}: ------------------------------->>>>>"))
  
  result <- as.list(df %>%
        select(df %>%
                 select_if(is.character) %>%
                 select(!contains("date", ignore.case = TRUE)) %>%
                 summarise(across(everything(), n_distinct)) %>%
                 select(where(~.x[[1]] < 40)) %>%
                 names()) %>%
        summarise(across(everything(), unique_values)))

  print(result)
  
  writeLines(
    stringr::str_glue("<<<<<--------------------------- :{df_name}"))
}

unique_char_col_vals("accepted_cc", accepted_cc)
```


Now read in the rejected credit card file:

```{r}
filename <- 
  paste0(data_path, "/rejected_credit_card_2007_to_2018Q4_2.csv")

rejected_cc <-
  read_csv(filename, progress = show_progress())
head(rejected_cc)
nrow(rejected_cc)
```
There are 1.3 million rows in the rejected file. But, since we used
grep to find rows, let's make sure the "credit_card" string is in the
"Loan Title" column, and nowhere else. Looking at the columns we have
today, it's probably not possible to put it anywhere else, but
if we run this script with a different file, there may be other
columns where "credit_card" will occur. Just good defensive programming.

```{r}
rejected_cc <- rejected_cc[rejected_cc$`Loan Title` == "credit_card", ]
nrow(rejected_cc)
```
There are still 1.3 million rows.

Replace names with lowercase, and replace spaces with underscores.

```{r}
names(rejected_cc) <- stringr::str_replace(
  stringr::str_to_lower(names(rejected_cc)), "\\s+", "_")
names(rejected_cc)
```

```{r}
rejected_cc %>%
  select(where(is.character)) %>%
  head()
```

```{r}
unique_char_col_vals("rejected_cc", rejected_cc)
```

## 1.2 Merging data and split into training, validation, test sets

Since the variables in the accepted and the rejected dataframes
are named differently, we will:
  
  * match variable names
  * create a new dataframe with values from accept/reject dataframes

### 1.2.1 Match variable names

Based on the variable names in the accepted and rejected
dataframes, the mapping between the names looks like:

rejected_cc            -> accepted_cc
-------------------------------------
"risk_score"           -> "fico_range_low", "fico_range_high"
"employment_length"    -> "emp_length"
"amount_requested"     -> "loan_amt"
"zip_code"             -> "zip_code"
"state"                -> "addr_state"
"debt-to-income_ratio" -> "dti"
"application_date"     -> "payment_plan_start_date" 

### 1.2.2 Create id for rejects

There is no id column for the reject dataframe, so generate one.
The only issue with a fake id here is that the same person may have
applied more than once. We will be considering all such applications
as coming from distinct individuals.

```{r}
typeof(accepted_cc$id)
sum(is.na(accepted_cc$id))
```

```{r}
min(rejected_cc$id)
max(rejected_cc$id)
```


```{r}
acc_rej_df <- tibble(
  "risk_score"
)
```

## 1.3 Selecting variable names

Now we will:

  * compare data between the two variables
  * decide if those two variables are useful to discriminate
    acceptance/rejection

We will start with the risk score:

```{r}
accepted_risk_scores <- 
  tibble(values = c(accepted_cc$fico_range_low,
                    accepted_cc$fico_range_high,
                    rejected_cc$risk_score),
         group = c(rep("accepted_fico_range_low",
                       length(accepted_cc$fico_range_low)),
                   rep("accepted_fico_range_high",
                       length(accepted_cc$fico_range_high)),
                   rep("rejected_risk_score",
                       length(rejected_cc$risk_score)))
  )

accepted_risk_scores %>%
  ggplot(aes(x = values, color = group)) +
  geom_freqpoly(binwidth = 30)
  # ggplot(aes(x = values, fill = group)) +
  # geom_histogram(bins = 50, position = "identity", alpha = 0.2)
```

As you can see, the rejected risk score histogram midpoint is where
the accepted fico range low values start. So this will be a good
variable to use for the acceptance criteria. We can use a cutoff
of the low value of the accepted fico range low histogram like this:
```{r}
score_threshold <- min(accepted_cc$fico_range_low)
score_threshold
```
Anything below this threshold can be rejected.
